depmix: An R-package for fitting mixture models on mixed multivariate data with Markov dependencies

### 2009-07-08 papers/jss/depmix-intro.Rnw
Moved around files in the papers directory
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@285 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\batchmode

@@

\author{Ingmar Visser\thanks{
Correspondence concerning this manual should be adressed to:
Ingmar Visser, Department of Psychology, University of Amsterdam,
Roetersstraat 15, 1018 WB, Amsterdam, The Netherlands}\\
Department of Psychology, University of Amsterdam\\
i.visser@uva.nl}

@@

\begin{abstract}

	\pkg{depmix} implements a general class of mixture models with
	Markovian dependencies between them in the R programming language
	\citep{R2007}.  This includes standard Markov models, latent/hidden
	Markov models, and latent class and finite mixture distribution models.
	The models can be fitted on mixed multivariate data with multinomial
	and/or gaussian distributions.  Parameters can be estimated subject to
	general linear constraints, and with optional inclusion of regression
	on (time-dependent) covariates.  Parameters estimation is done through
	a direct optimization approach with gradients using the \code{Rdonlp2}
	optimization routine.  A number of illustrative examples are included.

\end{abstract}

@@

Markov and latent Markov models are frequently used in the social
sciences, in different areas and applications.  In psychology, they
are used for modelling learning processes, see \citet{Wickens1982}, for an
overview, and \citet{Schmittmann2006} for a recent application.  In
economics, latent Markov models are commonly used as regime switching
models, see e.g.\ \citet{Kim1994} and \citet{Ghysels1994}.  Further
applications include speech recognition \citep{Rabiner1989}, EEG analysis
\citep{Rainer2000}, and genetics \citep{Krogh1998}.  In those latter areas
of application, latent Markov models are usually referred to as hidden
Markov models.

The \pkg{depmix} package was motivated by the fact that Markov models
are used commonly in the social sciences, but no comprehensive package
was available for fitting such models.  Common programs for Markovian
models include Panmark \citep{Panmark3-96}, and for latent class models
Latent Gold 
\citep{LatentGold3-2003}.  Those programs are lacking a number of
important features, besides not being freely available.  In
particular, \pkg{depmix}: 1) handles multiple case, or multiple group,
analysis; 2) handles arbitrarily long time series; 3) estimates models
with general linear constraints between parameters; 4) analyzes mixed
distributions, i.e., combinations of categorical and continuous
observed variables; 5) fits mixtures of latent Markov models to deal
with population heterogeneity; 6) can fit models with covariates.
Although \pkg{depmix} is specifically meant for dealing with
longitudinal or time series data, for say $T>100$, it can also handle
the limit case with $T=1$.  In those cases, there are no time
dependencies between observed data, and the model reduces to a finite
mixture model, or a latent class model.  In the next chapter, an outline
is provided of the model and the likelihood equations.  In the
chapters after that a number of examples are presented.

@@

@@

The latent Markov model is commonly associated with data of this type,
albeit usually only multinomial variables are considered.  However,
common estimation procedures, such as those implemented in
\citet{Panmark3-96} are not suitable for long time series due to
underflow problems.  In contrast, the hidden Markov model is typically
only used for `long' univariate time series.  In the next section, the
likelihood and estimation procedure for the hidden Markov model is
described, given data of the above form.

@@

The loglikelihood of hidden Markov models is usually computed by the
so-called forward-backward algorithm \citep{Bau66,Rab89}, or rather by
the forward part of this algorithm.  \cite{Lys2002} changed the forward
algorithm in such a way as to allow computing the gradients of the
loglikelihood at the same time.  They start by rewriting the
likelihood as follows (for ease of exposition the dependence on the
model parameters is dropped here):
\begin{equation}
	L_{T} = Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) = \prod_{t=1}^{T} 
Pr(\vc{O}_{t}|\vc{O}_{1}, 
	\ldots, \vc{O}_{t-1}), 
	\label{condLike}
\end{equation}
where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$. Note that for a 
simple, i.e.\ observed, Markov chain these probabilities reduce to 
$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots, 
\vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
The log-likelihood can now be expressed as:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log[Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1})].
	\label{eq:condLogl}
\end{equation}

To compute the log-likelihood, \cite{Lys2002} define the following 
(forward) recursion:
\begin{align}
	\phi_{1}(j) &:= Pr(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1}) 
	\label{eq:fwd1} \\
\begin{split}
	\phi_{t}(j) &:= Pr(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1}) \\
	&= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times 
(\Phi_{t-1})^{-1},
	\label{eq:fwdt} 
\end{split} 
\end{align}
where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$. Combining 
$\Phi_{t}=Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and 
equation~(\ref{eq:condLogl}) gives the following expression for the 
log-likelihood:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log \Phi_{t}.
	\label{eq:logl}
\end{equation}

@@

See equations 10--12  in \cite{Lys2002} for the score recursion 
functions of the hidden Markov model for a univariate time series. 
Here the corresponding score recursion for the multivariate mixture 
case are provided. The $t=1$ components of this score recursion are 
defined as (for an arbitrary parameter $\lambda_{1}$):
\begin{align}
\psi_{1}(j_{k};\lambda_{1}) &:=  \fpp Pr(\vc{O}_{1}|S_{1}=j_{k}) \\
\begin{split} 
	&= \left[  \fpp p_{k} \right] \pi_{j_{k}}b_{j_{k}}(\vc{O}_{1}) + 
	p_{k}\left[ \fpp \pi_{j_{k}} \right] b_{j_{k}}(\vc{O}_{1}) \\
	& \qquad  + p_{k}\pi_{j_{k}} \left[ \fpp 
b_{j_{k}}(\vc{O}_{1})\right],
\end{split} \label{eq:psi1}
\end{align}
and for $t>1$ the definition is:
\begin{align}
\psi_{t}(j_{k};\lambda_{1})  & =  \frac{\fpp Pr(\vc{O}_{1}, \ldots, 
\vc{O}_{t}, S_{t}=j_{k})}
			{Pr(\vc{O}_{1}, \ldots, \vc{O}_{t-1})}  \\
\begin{split} 
	& =  
			 \sum_{i=1}^{n_{k}} \Bigg\{ \psi_{t-1}(i;\lambda_{1})a_{ij_{k}} 
			 b_{j_{k}}(\vc{O}_{t}) \\ 
			 &\qquad +\phi_{t-1}(i) \left[ \fpp a_{ij_{k}} \right] b_{j_{k}} 
(\vc{O}_{t}) \\
			&\qquad +\phi_{t-1}(i)a_{ij_{k}}  \left[ \fpp b_{j_{k}} 
(\vc{O}_{t}) \right] \Bigg\} 
			\times (\Phi_{t-1})^{-1}.
\end{split} \label{eq:psit}
\end{align}

Using above equations, \cite{Lys2002} derive the following equation 
for the partial derivative of the likelihood:
\begin{equation}
	\fpp l_{T}= 	
		\frac{\mathbf{\Psi}_{T}(\lambda_{1})}{\mathbf{\Phi}_{T}},
\end{equation}
where $\Psi_{t}=\sum_{k=1}^{K} \sum_{i=1}^{n_{k}} 
\psi_{t}(j_{k};\lambda_{1})$. 
Starting from the equation from the logarithm of the likelihood, this 
is easily seen to be correct: 
\begin{eqnarray*}
	\fpp \log Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) &=& Pr(\vc{O}_{1}, 
\ldots, \vc{O}_{T})^{-1} 
	\fpp Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) \\
	&=&  \frac{Pr(\vc{O}_{1}, \ldots, \vc{O}_{T-1})}{Pr(\vc{O}_{1}, 
\ldots, \vc{O}_{T})}  \Psi_{T} (\lambda_{1}) \\
	&=&  \frac{\mathbf{\Psi}_{T}(\lambda_{1})}{\mathbf{\Phi}_{T}}.
\end{eqnarray*}

@@

Constraining and fixing parameters is done in a similar fashion as the
\code{pa} command that is used in LISREL \citep{Lis99}. The
\code{conpat} argument to the \code{fitdmm}-function specifies for
each parameter in the model whether it's fixed (0) or free (1 or
higher). Equality constraints can be imposed by having two parameters have 
the same number in the \code{conpat} vector. When only fixed values are
required the \code{fixed} argument can be used instead of \code{conpat},
with zeroes for fixed parameters and other values (ones e.g.) for non-fixed
parameters.

@@

\nocite{Raijmakers2001}
\nocite{Maas2005b}
\nocite{Schmittmann2005b}
\nocite{McCutcheon1987}

\bibliography{all}

>>>>

%\batchmode

@@

\author{Version: 0.9.1 \\ Ingmar Visser\thanks{Correspondence concerning this manual
should be adressed to: Ingmar Visser, Department of Psychology,
University of Amsterdam, Roetersstraat 15, 1018 WB, Amsterdam, The
Netherlands}\\
Developmental Processes Research Group\\
Department of Psychology, University of Amsterdam\\
i.visser@uva.nl}

@@

\begin{abstract}
	\pkg{depmix} implements a general class of mixture models with
	Markovian dependencies between them in the R programming language
	\citep{R2006}.  This includes standard Markov models,
	latent/hidden Markov models, and latent class and finite mixture
	distribution models.  The models can be fitted on mixed
	multivariate data from a number of distributions including the
	binomial, multinomial, gaussian, lognormal and Weibull
	distributions.  Parameters can be estimated subject to general
	linear constraints, and with optional inclusion of regression on
	(time-dependent) covariates.  Parameter estimation is done
	through a direct optimization approach with gradients using the
	\code{nlm} optimization routine.  Optional support for using the NPSOL
	optimization routines is provided as well.  A number of illustrative
	examples are included.
\end{abstract}

@@

Markov and latent Markov models are frequently used in the social
sciences, in different areas and applications.  In psychology, they
are used for modelling learning processes, see \citet{Wickens1982}, for an
overview, and \citet{Schmittmann2005} for a recent application.  In
economics, latent Markov models are commonly used as regime switching
models, see e.g.\ \citet{Kim1994} and \citet{Ghysels1994}.  Further
applications include speech recognition \citep{Rabiner1989}, EEG analysis
\citep{Rainer2000}, and genetics \citep{Krogh1998}.  In those latter areas
of application, latent Markov models are usually referred to as hidden
Markov models.

The \pkg{depmix} package was motivated by the fact that Markov models
are used commonly in the social sciences, but no comprehensive package
was available for fitting such models.  Common programs for Markovian
models include Panmark \citep{Pol1996}, and for latent class models
Latent Gold 
\citep{Vermunt2003}.  Those programs are lacking a number of
important features, besides not being freely available.  In
particular, \pkg{depmix}: 1) handles multiple case, or multiple group,
analysis; 2) handles arbitrarily long time series; 3) estimates models
with general linear constraints between parameters; 4) analyzes mixed
distributions, i.e., combinations of categorical and continuous
observed variables; 5) fits mixtures of latent Markov models to deal
with population heterogeneity; 6) can fit models with covariates.
Although \pkg{depmix} is specifically meant for dealing with
longitudinal or time series data, for say $T>100$, it can also handle
the limit case with $T=1$.  In those cases, there are no time
dependencies between observed data, and the model reduces to a finite
mixture model, or a latent class model.  In the next chapter, an outline
is provided of the model and the likelihood equations.  In the
chapters after that a number of examples are presented.

@@

Development of this pacakge was supported by European Commission grant
51652 (NEST) and by a VENI grant from the Dutch Organization for Scientific
Research (NWO) to Ingmar Visser.

@@

The latent Markov model is commonly associated with data of this type,
albeit usually only multinomial variables are considered.  However,
common estimation procedures, such as those implemented in
\citet{Pol1996} are not suitable for long time series due to
underflow problems.  In contrast, the hidden Markov model is typically
only used for `long' univariate time series.  In the next section, the
likelihood and estimation procedure for the hidden Markov model is
described, given data of the above form.

@@

The loglikelihood of hidden Markov models is usually computed by the
so-called forward-backward algorithm \citep{Baum1966,Rabiner1989}, or rather by
the forward part of this algorithm.  \cite{Lystig2002} changed the forward
algorithm in such a way as to allow computing the gradients of the
loglikelihood at the same time.  They start by rewriting the
likelihood as follows (for ease of exposition the dependence on the
model parameters is dropped here):
\begin{equation}
	L_{T} = Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) = \prod_{t=1}^{T} 
Pr(\vc{O}_{t}|\vc{O}_{1}, 
	\ldots, \vc{O}_{t-1}), 
	\label{condLike}
\end{equation}
where $Pr(\vc{O}_{1}|\vc{O}_{0}):=Pr(\vc{O}_{1})$. Note that for a 
simple, i.e.\ observed, Markov chain these probabilities reduce to 
$Pr(\vc{O}_{t}|\vc{O}_{1},\ldots, 
\vc{O}_{t-1})=Pr(\vc{O}_{t}|\vc{O}_{t-1})$.
The log-likelihood can now be expressed as:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log[Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1})].
	\label{eq:condLogl}
\end{equation}

To compute the log-likelihood, \cite{Lystig2002} define the following 
(forward) recursion:
\begin{align}
	\phi_{1}(j) &:= Pr(\vc{O}_{1}, S_{1}=j) = \pi_{j} b_{j}(\vc{O}_{1}) 
	\label{eq:fwd1} \\
\begin{split}
	\phi_{t}(j) &:= Pr(\vc{O}_{t}, S_{t}=j|\vc{O}_{1}, \ldots, 
\vc{O}_{t-1}) \\
	&= \sum_{i=1}^{N} [\phi_{t-1}(i)a_{ij}b_{j}(\vc{O}_{t})] \times 
(\Phi_{t-1})^{-1},
	\label{eq:fwdt} 
\end{split} 
\end{align}
where $\Phi_{t}=\sum_{i=1}^{N} \phi_{t}(i)$. Combining 
$\Phi_{t}=Pr(\vc{O}_{t}|\vc{O}_{1}, \ldots, \vc{O}_{t-1})$, and 
equation~(\ref{eq:condLogl}) gives the following expression for the 
log-likelihood:
\begin{equation}
	l_{T} = \sum_{t=1}^{T} \log \Phi_{t}.
	\label{eq:logl}
\end{equation}

@@

See equations 10--12  in \cite{Lystig2002} for the score recursion 
functions of the hidden Markov model for a univariate time series. 
Here the corresponding score recursion for the multivariate mixture 
case are provided. The $t=1$ components of this score recursion are 
defined as (for an arbitrary parameter $\lambda_{1}$):
\begin{align}
\psi_{1}(j_{k};\lambda_{1}) &:=  \fpp Pr(\vc{O}_{1}|S_{1}=j_{k}) \\
\begin{split} 
	&= \left[  \fpp p_{k} \right] \pi_{j_{k}}b_{j_{k}}(\vc{O}_{1}) + 
	p_{k}\left[ \fpp \pi_{j_{k}} \right] b_{j_{k}}(\vc{O}_{1}) \\
	& \qquad  + p_{k}\pi_{j_{k}} \left[ \fpp 
b_{j_{k}}(\vc{O}_{1})\right],
\end{split} \label{eq:psi1}
\end{align}
and for $t>1$ the definition is:
\begin{align}
\psi_{t}(j_{k};\lambda_{1})  & =  \frac{\fpp Pr(\vc{O}_{1}, \ldots, 
\vc{O}_{t}, S_{t}=j_{k})}
			{Pr(\vc{O}_{1}, \ldots, \vc{O}_{t-1})}  \\
\begin{split} 
	& =  
			 \sum_{i=1}^{n_{k}} \Bigg\{ \psi_{t-1}(i;\lambda_{1})a_{ij_{k}} 
			 b_{j_{k}}(\vc{O}_{t}) \\ 
			 &\qquad +\phi_{t-1}(i) \left[ \fpp a_{ij_{k}} \right] b_{j_{k}} 
(\vc{O}_{t}) \\
			&\qquad +\phi_{t-1}(i)a_{ij_{k}}  \left[ \fpp b_{j_{k}} 
(\vc{O}_{t}) \right] \Bigg\} 
			\times (\Phi_{t-1})^{-1}.
\end{split} \label{eq:psit}
\end{align}

Using above equations, \cite{Lystig2002} derive the following equation 
for the partial derivative of the likelihood:
\begin{equation}
	\fpp l_{T}= 	
		\frac{\mathbf{\Psi}_{T}(\lambda_{1})}{\mathbf{\Phi}_{T}},
\end{equation}
where $\Psi_{t}=\sum_{k=1}^{K} \sum_{i=1}^{n_{k}} 
\psi_{t}(j_{k};\lambda_{1})$. 
Starting from the equation from the logarithm of the likelihood, this 
is easily seen to be correct: 
\begin{eqnarray*}
	\fpp \log Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) &=& Pr(\vc{O}_{1}, 
\ldots, \vc{O}_{T})^{-1} 
	\fpp Pr(\vc{O}_{1}, \ldots, \vc{O}_{T}) \\
	&=&  \frac{Pr(\vc{O}_{1}, \ldots, \vc{O}_{T-1})}{Pr(\vc{O}_{1}, 
\ldots, \vc{O}_{T})}  \Psi_{T} (\lambda_{1}) \\
	&=&  \frac{\mathbf{\Psi}_{T}(\lambda_{1})}{\mathbf{\Phi}_{T}}.
\end{eqnarray*}

@@

Constraining and fixing parameters is done in a similar fashion as the
\code{pa} command that is used in LISREL \citep{Joreskog1999}. The
\code{conpat} argument to the \code{fitdmm}-function specifies for
each parameter in the model whether it's fixed (0) or free (1 or
higher). Equality constraints can be imposed by having two parameters have 
the same number in the \code{conpat} vector. When only fixed values are
required the \code{fixed} argument can be used instead of \code{conpat},
with zeroes for fixed parameters and other values (ones e.g.) for non-fixed
parameters.

@@

\section{Known issues and future plans}

\paragraph{Constraint violations} Parameter optimization is done by adding
a penalty to the log likelihood whenever constraints are not satisfied, ie
linear inequality and box constraints.  Equality constraints are fitted by
reparametrization and do not suffer from this problem.  Refer to the
section on parameter estimation and the \code{fitdmm} help page on how to
deal with constraint violations. Future plans include using Lagrange
multipliers to overcome this problem.

\nocite{Raijmakers2001}
\nocite{Maas2005b}
\nocite{Schmittmann2006}
\nocite{McCutcheon1987}
\nocite{Visser2006b}

\bibliography{all,ingmar}
